### 2016-08-16 latex/input/abstract.tex
paper draft v1.1: most minor edits completed

The recent advent of Metagenome-Wide Association Studies (MGWAS) has allowed for increased accuracy in the prediction of patient phenotype (disease), but has also presented big data challenges. Meanwhile, Multiple Instance Learning (MIL) is useful in the domain of bioinformatics because, in addition to classifying patient phenotype, it can also identify individual parts of the microbiome that are indicative of that phenotype, leading to better understanding of the disease. We demonstrate a novel, efficient, and effective MIL-based computational pipeline to predict patient phenotype from MGWAS data. Specifically, we use a Distance-based Bag of Words method, which has been shown to be one of the most effective and efficient MIL methods. This involves assembly of the metagenomic data, clustering of the assembled contigs, extracting features from the contigs, and using a standard SVM classifier to predict patient labels and identify the most relevant read clusters. With the exception of the given labels for the patients, this entire process is \emph{de novo} (unsupervised). We use data from a well-known MGWAS study of patients with Type-2 Diabetes and show that our pipeline significantly outperforms the classifier used in that paper, as well as other common MIL methods. We call our pipeline "CAMIL", which stands for Clustering and Assembly with Multiple Instance Learning.

>>>>

The recent advent of Metagenome-Wide Association Studies (MGWAS) has allowed for increased accuracy in the prediction of patient phenotype (disease), but has also presented big data challenges. Meanwhile, Multiple Instance Learning (MIL) is useful in the domain of bioinformatics because, in addition to classifying patient phenotype, it can also identify individual parts of the microbiome that are indicative of that phenotype, leading to better understanding of the disease. We demonstrate a novel, efficient, and effective MIL-based computational pipeline to predict patient phenotype from MGWAS data. Specifically, we use a Distance-based Bag of Words method, which has been shown to be one of the most effective and efficient MIL methods. This involves assembly of the metagenomic sequence data, clustering of the assembled contigs, extracting features from the contigs, and using an SVM classifier to predict patient labels and identify the most relevant read clusters. With the exception of the given labels for the patients, this entire process is \emph{de novo} (unsupervised). We use data from a well-known MGWAS study of patients with Type-2 Diabetes and show that our pipeline significantly outperforms the classifier used in that paper, as well as other common MIL methods. We call our pipeline ``CAMIL", which stands for Clustering and Assembly with Multiple Instance Learning.

### 2016-08-16 latex/input/experiments.tex
paper draft v1.1: most minor edits completed

The 218 publicly-available and labeled patients were downloaded from NCBI and converted to FASTQ format using the SRA toolkit, and labeled according to the labels on EBI, as described in the Materials section. The data was split evenly into training and test sets, with each containing 109 patients. The training set had 54 T2D patients and 55 control patients, while the test set had 53 T2D patients and 56 control patients. Each read was assembled with SOAPdenovo2, with the k-mer length set to 51, reads cut off after 100 base pairs (original length of 180 base pairs), and the average insert size set to 350 in accordance with the reported average insert size reported in the MGWAS study \cite{qin041012}. Assembly for each file took 7-30 minutes, depending on the file size, and each patient was assembled in parallel. Combining the files into one file took 12 minutes. Assembly was used for all of the classification methods tested, because the combining of individual reads and reduction in total data size made classification feasible.

@@

We implemented two different versions of the pipeline: one that uses D-BoW feature extraction, and one that uses H-BoW feature extraction. These results are denoted as "CAMIL D-BoW" and "CAMIL H-BoW", respectively, in the results tables and graphs. Clustering for the pipeline with UCLUST took 10 hours and 51 minutes. Without the assembly step reducing the size of the data, clustering would not have been feasible. Feature extraction and SVM-light classification took 10 minutes and 17 seconds to run for D-BoW, versus 5 minutes and 14 seconds for H-BoW.

@@

MISVM \cite{andrews02} and sbMIL \cite{bunescu07} are two of the classic Multiple Instance Learning algorithms that fall into what Amores calls the "Instance Space" (IS) methods, in that they only use "local" information based on comparisons between individual instances and treat bag labels as aggregations of instance labels \cite{amores13}. Additionally, both of these methods follow the standard MIL assumption that bags with negative labels contain only negative instances, whereas positive bags contain one or more positive instances \cite{amores13}. sbMIL specifically assumes that positive bags contain few positive instances \cite{bunescu07}. We include these algorithms as an example of many of the early MIL algorithms, which usually fell into the IS paradigm and used the standard MIL assumption. For the implementation of these methods, we used an open-source Python implementation by Doran \cite{doran14}, which is available on GitHub\footnote{https://github.com/garydoranjr/misvm}.

@@

The Group-Instance Cost Function (GICF) is a method proposed by Kotzias et al. that learns instance labels in addition to group labels \cite{kotzias15}. The cost function uses a kernel that measures similarity between instances and a penalty on the difference between instance labels to generate instance labels \cite{kotzias15}. It then sets the group label to be the average instance label of all instances in that group, using a penalty on the difference between the predicted group label and the actual group label \cite{kotzias15}. Ideally, this would cause instances that are similar to each other to have similar labels, and predicted group labels to correspond closely to reality. Unlike MISVM and sbMIL, GICF explicitly does not hold the standard MIL assumption, instead favoring the collective assumption. However, because this method compares only individual instances and not entire bags, and treats bag labels simply as aggregations of instance labels, GICF is still an Instance Space method.

@@

The methods in the original paper are neither MIL-based, nor are they entirely de novo apart from patient labels. The authors first performed de novo assembly with SOAPdenovo2 \cite{luo12} and then use a tool called MetaGeneMark \cite{zhu10, besemer99} for de novo prediction of genes from the assembled contigs \cite{qin041012}. They then combined these genes with an existing gene catalog, MetaHIT \cite{qin030410}, and carried out taxonomic assignment and functional annotation of the genes using the KEGG \cite{kanehisa00} and eggNOG \cite{powell12} databases, as well as 2,890 other reference genomes \cite{qin041012}. The authors defined gene markers by mapping the sequence reads from the MGWAS dataset to the updated gene catalog. They identified the 50 most important gene markers with the minimum redundancy - maximum relevance (mRMR) \cite{peng05} method, using the "sideChannelAttack" R package and then used these 50 gene markers for SVM classification of T2D phenotype, using the "e1071" R package for the SVM \cite{qin041012}. Thus, the method in the original paper first applies de novo assembly and gene prediction methods, but then uses a number of references to identify the gene markers to be used in classification. From their results, the authors generated an Area Under Curve - Receiver Operating Characteristic (AUC-ROC) graph. We manually computed the accuracy and F1 score with an ideally-chosen decision boundary based on the results the authors provided in their supplementary tables.

@@

\begin{table}[h]
\begin{center}
\caption{Comparison of the performance of various methods on the MGWAS dataset. 
\label{tab:comp}}
\begin{tabular}{|c|ccc|}\hline
Method & Accuracy & F1-Score & AUC-ROC\\\hline
MISVM & 50.00 & 50.00 & 50.00\\\hline
sbMIL & 50.00 & 50.00 & 50.00\\\hline
GICF & 60.00 & 71.43 & 66.00\\\hline
Original & 82.61 & 80.02 & 81.00\\\hline 
CAMIL H-BoW & 91.74 & 91.43 & 95.00\\\hline
CAMIL D-BoW & \bf{92.66} & \bf{92.31} & \bf{96.00}\\\hline
\end{tabular}
\end{center}
\end{table}

@@

\subsection{Deriving Instance "Labels"}

@@

One of the benefits of using Multiple Instance Learning methods is that we can attempt to discover instance "labels". In fact, we did not attempt to apply static, unchanging labels to individual reads or clusters, since organisms are affected by their interactions with each other. For instance, a patient with X amount of microbe A or X amount of microbe B may be healthy, but with X/2 amount of microbe A and X/2 amount of microbe B they may be sick. This is illustrated in Figure \ref{instance-labels}. This example is very simplified, but it explains why static instance labels are insufficient.

However, we can infer from the SVM decision boundary which clusters appear to be most relevant to the disease diagnosis. Since feature vectors are multiplied by the weight vector of the decision boundary to determine the label of the patient, we can assume that clusters with the highest weights in the weight vector are most relevant to the disease diagnosis. For instance, if the Ith scalar in the weight vector is has the highest value of any of the weights, then cluster I is likely to play a major role in the disease pathology. Similarly, the most negative weights in the weight vector indicate clusters whose presence in a patient indicates that they likely do not have the disease. Because the data is metagenomic, the clusters represent both phylogenetic and functional similarity, so identifying the most relevant clusters can help discover more about the pathology of the disease. For Type 2 Diabetes, which is a complex phenotype and a disease that is both common and deadly, this is potentially quite valuable.

In the original MGWAS paper, the authors identify 50 important gene markers with mRMR that are used for their classifier. Conversely, CAMIL uses all of the data to train and test the classifier, and subsequently identifies significant clusters based on the classification results. We printed the 25 highest-weighted and 25 lowest-weighted clusters, out of 3188 total clusters. The highest-weighted cluster had a weight of 0.012109, while the lowest-weighted cluster had a weight of -0.003668. The average of the 25 highest cluster weights was 0.010402, while it was -0.002753 for the lowest 25. The average of all 3188 cluster weights was 0.001003, and the median was 0.000254. These results indicate that the highest weights were much larger than the lowest weights and skewed the average weight to be much higher than the median weight. Intuitively, this appears to make sense, as there should be a relatively small number of key clusters whose presence is actually indicative of type 2 diabetes, while most other clusters aren't particularly relevant in this case and whose weights are just noise. Thus, the weights obtained by this method appear to be plausible. In contrast, the labels obtained by GICF were barely differentiated from each other at all. The instance weights determined by CAMIL are shown in Figure \ref{instance-scatter}.

>>>>

The 218 publicly-available and labeled patients were downloaded from NCBI and converted to FASTQ format using the SRA toolkit, and labeled according to the labels on EBI, as described in the Materials section. The data was split evenly into training and test sets, with each containing 109 patients. The training set had 54 T2D patients and 55 control patients, while the test set had 53 T2D patients and 56 control patients. Each read was assembled with SOAPdenovo2, with the k-mer length set to 51, reads cut off after 100 base pairs (original length of 180 base pairs), and the average insert size set to 350 in accordance with the reported average insert size reported in the MGWAS study \cite{qin041012}. Assembly for each file took 7-30 minutes, depending on the file size. The assembly of the patient files is embarrassingly parallel, so the assembly of each file could be done at the same time. Combining the files into one file took 12 minutes. Assembly was used for all of the classification methods tested, because the combining of individual reads and reduction in total data size made classification feasible.

@@

We implemented two different versions of the pipeline: one that uses D-BoW feature extraction, and one that uses H-BoW feature extraction. These results are denoted as ``CAMIL D-BoW" and ``CAMIL H-BoW", respectively, in the results tables and graphs. Clustering for the pipeline with UCLUST took 10 hours and 51 minutes. Without the assembly step reducing the size of the data, clustering would not have been feasible. Feature extraction and SVM-light classification took 10 minutes and 17 seconds to run for D-BoW, versus 5 minutes and 14 seconds for H-BoW.

@@

MISVM \cite{andrews02} and sbMIL \cite{bunescu07} are two of the classic Multiple Instance Learning algorithms that fall into what Amores calls the ``Instance Space" (IS) methods, in that they only use ``local" information based on comparisons between individual instances and treat bag labels as aggregations of instance labels \cite{amores13}. Additionally, both of these methods follow the standard MIL assumption that bags with negative labels contain only negative instances, whereas positive bags contain one or more positive instances \cite{amores13}. sbMIL specifically assumes that positive bags contain few positive instances \cite{bunescu07}. We include these algorithms as an example of many of the early MIL algorithms, which usually fell into the IS paradigm and used the standard MIL assumption. For the implementation of these methods, we used an open-source Python implementation by Doran \cite{doran14}, which is available on GitHub\footnote{https://github.com/garydoranjr/misvm}.

@@

The Group-Instance Cost Function (GICF) is a method proposed by Kotzias et al. that learns instance labels in addition to group labels \cite{kotzias15}. The cost function uses a kernel that measures similarity between instances and a penalty on the difference between instance labels to generate instance labels \cite{kotzias15}. It then sets the group label to be the average instance label of all instances in that group, using a penalty on the difference between the predicted group label and the actual group label \cite{kotzias15}. Ideally, this would cause instances that are similar to each other to have similar predicted labels, and predicted group labels to correspond closely to reality. Unlike MISVM and sbMIL, GICF explicitly does not hold the standard MIL assumption, instead favoring the collective assumption. However, because this method compares only individual instances and not entire bags, and treats bag labels simply as aggregations of instance labels, GICF is still an Instance Space method.

@@

The methods in the original paper are neither MIL-based, nor are they entirely de novo apart from patient labels. The authors first performed de novo assembly with SOAPdenovo2 \cite{luo12} and then use a tool called MetaGeneMark \cite{zhu10, besemer99} for de novo prediction of genes from the assembled contigs \cite{qin041012}. They then combined these genes with an existing gene catalog, MetaHIT \cite{qin030410}, and carried out taxonomic assignment and functional annotation of the genes using the KEGG \cite{kanehisa00} and eggNOG \cite{powell12} databases, as well as 2,890 other reference genomes \cite{qin041012}. The authors defined gene markers by mapping the sequence reads from the MGWAS dataset to the updated gene catalog. They identified the 50 most important gene markers with the minimum redundancy - maximum relevance (mRMR) \cite{peng05} method, using the ``sideChannelAttack" R package and then used these 50 gene markers for SVM classification of T2D phenotype, using the ``e1071" R package for the SVM \cite{qin041012}. Thus, the method in the original paper first applies de novo assembly and gene prediction methods, but then uses a number of references to identify the gene markers to be used in classification. From their results, the authors generated an Area Under Curve - Receiver Operating Characteristic (AUC-ROC) graph. We manually computed the accuracy and F1 score with an ideally-chosen decision boundary based on the results the authors provided in their supplementary tables.

@@

\begin{table}[h]
\begin{center}
\caption{Comparison of the performance of various methods on the MGWAS dataset. 
\label{tab:comp}}
\begin{tabular}{|c|ccc|}\hline
Method & Accuracy & F1-Score & AUC-ROC\\\hline
MISVM & 50.00 & 50.00 & 50.00\\\hline
sbMIL & 50.00 & 50.00 & 50.00\\\hline
GICF & 60.00 & 71.43 & 66.00\\\hline
mRMR + SVM & 82.61 & 80.02 & 81.00\\\hline 
CAMIL H-BoW & 91.74 & 91.43 & 95.00\\\hline
CAMIL D-BoW & \bf{92.66} & \bf{92.31} & \bf{96.00}\\\hline
\end{tabular}
\end{center}
\end{table}

@@

\subsection{Deriving Instance ``Labels"}

@@

One of the benefits of using Multiple Instance Learning methods is that we can attempt to discover instance ``labels". In fact, we did not attempt to apply static, unchanging labels to individual reads or clusters, since organisms are affected by their interactions with each other. For instance, a patient with X amount of microbe A or X amount of microbe B may be healthy, but with X/2 amount of microbe A and X/2 amount of microbe B they may be sick. This is illustrated in Figure \ref{instance-labels}. This example is simple, but it explains why static instance labels are insufficient.

However, we can infer from the SVM decision boundary which clusters appear to be most relevant to the disease diagnosis. Since feature vectors are multiplied by the weight vector of the decision boundary to determine the label of the patient, we can assume that clusters with the highest weights in the weight vector are most relevant to the disease diagnosis. For instance, if the \emph{i}th scalar in the weight vector is has the highest value of any of the weights, then cluster \emph{i} is likely to play a major role in the disease pathology. Similarly, the most negative weights in the weight vector indicate clusters whose presence in a patient indicates that they likely do not have the disease. Because the data is metagenomic, the clusters represent both phylogenetic and functional similarity, so identifying the most relevant clusters can help discover more about the pathology of the disease. For Type 2 Diabetes, which is a complex phenotype and a disease that is both common and deadly, this is potentially quite valuable.

In the original MGWAS paper, the authors identify 50 important gene markers with mRMR that are used for their classifier. Conversely, CAMIL uses all of the data to train and test the classifier, and subsequently identifies significant clusters based on the classification results. We printed the 25 highest-weighted and 25 lowest-weighted clusters, out of 3188 total clusters. The highest-weighted cluster had a weight of 0.012109, while the lowest-weighted cluster had a weight of -0.003668. The average of the 25 highest cluster weights was 0.010402, while it was -0.002753 for the lowest 25. The average of all 3188 cluster weights was 0.001003, and the median was 0.000254. These results indicate that the highest weights were much larger than the lowest weights and skewed the average weight to be much higher than the median weight. Intuitively, this appears to make sense, as there should be a relatively small number of key clusters whose presence is actually indicative of type 2 diabetes, while most other clusters are not particularly relevant in this case and whose weights are just noise. Thus, the weights obtained by this method appear to be plausible. In contrast, the labels obtained by GICF were barely differentiated from each other at all. The instance weights determined by CAMIL are shown in Figure \ref{instance-scatter}.

### 2016-08-16 latex/input/intro.tex
paper draft v1.1: most minor edits completed

Metagenomics has several advantages: (i) microbes are now understood to be the underlying cause of many human diseases and are also critical to many chemical processes and overall health; (ii) it is believed that most microbes have not been laboratory-cultured and thus remain unknown; (iii) whereas other methods such as 16S rRNA analysis are mainly useful for predicting the species of microbes (\emph{phylogeny}), metagenomics contains other critical information from the microbial genomes that determine how these microbes function and affect diseases and chemical processes (\emph{functional} information) \cite{handelsman04}. Summarily, metagenomics allows us to view microbial data that is not accessible to us via traditional laboratory culturing and allows for both phylogenetic and functional profiling of those microbes. Thus, studying microbial metagenomics is an effective way to predict and model human disease, also known as clinical \emph{phenotype}.

In this study, we develop an efficient classifier that predicts whether or not a patient has a disease based on their microbiome. This can be viewed as a Multiple Instance Learning (MIL) problem, in which we have several \emph{bags} of instances, and we have labels for the bags, but not for each instance within them. In this case, we have a patient (bag) and a label for each patient (whether or not they have a disease), but no labels for each patient's sequence reads (instances). Specifically, we use a Distance-based Bag of Words (D-BoW) method, discussed further in the Background section, which has been shown to be among the most effective and efficient MIL methods \cite{amores13}. We also show that our pipeline can be used for Histogram-based Bag of Words (H-BoW) methods. MIL has been studied in many contexts, but it has rarely if ever been studied in the context of predicting clinical phenotype based on metagenomic data from the microbiome. However, since datasets in this domain frequently have patient-level labels but almost never have instance-level labels, this is a well-suited domain for multiple instance learning.

We used data from a Metagenome-Wide Association Study (MGWAS), which compares microbial metagenomic data between many patients with or without a given phenotype. Because MGWAS studies contain many expert-labeled patients and the metagenomic data associated with those patients, they are very useful for phenotype prediction but also cause many computational challenges. The data is very large (multiple terabytes) and high-dimensional (thousands of dimensions). Additionally, due to the nature of shotgun sequencing, most of the reads are not useful by themselves, and must first be assembled. \emph{Assembly} is the process of finding pairs of reads in which the end of one read overlaps with the beginning of another, signifying that they are probably contiguous reads from the same genome, and combining them. This process is repeated as much as possible to form long strings called \emph{contigs}. Assembly also reduces the size and dimensionality of the data by discarding reads that cannot be assembled successfully. Most of the discarded reads would not be particularly useful anyway, since the assembly process could not establish their relationship to other reads and many of them are due to sequencing machine errors. The reduction in data size also allows for clustering, which is not feasible for massive datasets. Clustering uses string similarity measures to group similar reads into "clusters", which is a way of identifying which species of microbe each read corresponds to. The alternative, "aligning" the reads with known genomes, is both time-inefficient and impractical for metagenomics, since many of the involved microbes have not yet had their genomes sequenced. From the clustering output, we extract feature vectors, which are then fed into a standard Support Vector Machine (SVM) classifier.

Thus, the entire pipeline consists of assembling the reads of each patient, combining the resulting contigs from each patient into a single file, clustering the contigs, extracting features from the clustering output, and performing classification with the SVM. This process is explained in further detail in the methods section. We refer to the pipeline as "CAMIL", which stands for Clustering and Assembly with Multiple Instance Learning. We then compare the results of our classifier against the classifier used in the MGWAS study from which we derived our data. We show that our classifier shows significantly improved performance. This is discussed further in the Experimental Results section.

>>>>

Metagenomics has several advantages: (i) microbes are now understood to be the underlying cause of many human diseases and are also critical to many chemical processes and overall health \cite{handelsman04}; (ii) it is believed that most microbes have not been laboratory-cultured and thus remain unknown \cite{handelsman04}; (iii) whereas other methods such as 16S rRNA analysis are mainly useful for predicting the species of microbes (\emph{phylogeny}), metagenomics contains other critical information from the microbial genomes that determine how these microbes function and affect diseases and chemical processes (\emph{functional} information) \cite{handelsman04}. Summarily, metagenomics allows us to view microbial data that is not accessible to us via traditional laboratory culturing and allows for both phylogenetic and functional profiling of those microbes. Thus, studying microbial metagenomics is an effective way to predict and model human disease, also known as clinical \emph{phenotype}.

In this study, we develop an efficient classifier that predicts whether or not a patient has a disease based on their microbiome. We view this as a Multiple Instance Learning (MIL) problem, in which we have several \emph{bags} of instances, and we have labels for the bags, but not for each instance within them. In this case, we have a patient (bag) and a label for each patient (whether or not they have a disease), but no labels for each patient's sequence reads (instances). Specifically, we use a Distance-based Bag of Words (D-BoW) method, discussed further in the Background section, which has been shown to be among the most effective and efficient MIL methods \cite{amores13}. We also show that our pipeline can be used for Histogram-based Bag of Words (H-BoW) methods. MIL has been studied in many contexts, but it has rarely if ever been studied in the context of predicting clinical phenotype based on metagenomic data from the microbiome. However, since datasets in this domain frequently have patient-level labels but almost never have instance-level labels, this is a well-suited domain for multiple instance learning.

We used data from a Metagenome-Wide Association Study (MGWAS), which compares microbial metagenomic data between many patients with or without a given phenotype. MGWAS studies contain many expert-labeled patients and the metagenomic data associated with those patients, so they are useful for phenotype prediction but also cause many computational challenges. The data is very large (multiple terabytes) and high-dimensional (thousands of dimensions). Additionally, due to the nature of shotgun sequencing, most of the reads are not useful by themselves, and must first be assembled. \emph{Assembly} is the process of combining pairs of reads in which the end of one read overlaps with the beginning of another, signifying that they are probably contiguous reads from the same genome. This process is repeated as much as possible to form long strings called \emph{contigs}. Assembly also reduces the size and dimensionality of the data by discarding reads that cannot be assembled successfully. The reduction in data size also allows for clustering, which is not feasible for massive datasets. \emph{Clustering} uses string similarity measures to group similar reads into ``clusters", which is a way of identifying which species of microbe each read corresponds to. The alternative, ``aligning" the reads with known genomes, is both time-inefficient and impractical for metagenomics, since many of the involved microbes have not yet had their genomes sequenced. From the clustering output, we extract feature vectors, which are then fed into a Support Vector Machine (SVM) classifier.

Thus, the entire pipeline consists of assembling the reads of each patient, combining the resulting contigs from each patient into a single file, clustering the contigs, extracting features from the clustering output, and performing classification with the SVM. This process is explained in further detail in the methods section. We refer to the pipeline as ``CAMIL", which stands for Clustering and Assembly with Multiple Instance Learning. We then compare the results of our classifier against the classifier used in the MGWAS study from which we derived our data. We show that our classifier shows significantly improved performance. This is discussed further in the Experimental Results section.

### 2016-08-16 latex/input/materials.tex
paper draft v1.1: most minor edits completed

We used data from a well-known Metagenome-Wide Association Study by Qin et al. of Type 2 Diabetes (T2D) in Chinese patients \cite{qin041012}. This study was chosen because it is one of the only MGWAS studies that made its data available online and labeled the phenotype of the patients, and is the largest among those studies. The full dataset used in this study contains 368 patients \cite{qin041012}, but only 218 of them were labeled and available online. Each patient file was downloaded from NCBI\footnote{http://trace.ncbi.nlm.nih.gov/Traces/study/?acc=SRP011011}
and converted to FASTQ format using the SRA toolkit\footnote{http://www.ncbi.nlm.nih.gov/Traces/sra/sra.cgi?view=software}.
The labels were found on the EBI website\footnote{https://www.ebi.ac.uk/metagenomics/projects/SRP011011}. The total size of these 218 FASTQ files was 2.3 terabytes, with an average size of 10.82 gigabytes per patient file. Qin et al. develop a simple T2D classifier using a minimum redundancy–maximum relevance (mRMR) method \cite{peng05} for feature selection and an SVM for classification, based on the R packages "sideChannelAttack" and "1071", respectively \cite{qin041012}. They achieved an AUC-ROC of 0.81 by training a classifier on 345 of the patients and using the remaining 23 as a test set \cite{qin041012}. While we did not have access to all of the same patients that they had, we were able to replicate their method and test it on the subset of data that was publicly available. The study also called for more extensive testing of gut microbiota classifiers \cite{qin041012}.

\subsection{Software and Hardware Details}
We used the ARGO computing cluster available at George Mason University\footnote{http://orc.gmu.edu/research-computing/argo-cluster/argo-hardware-specs/}. The clustering and classification phases were run on one of the compute nodes available on the cluster. The cluster is configured with 35 Dell C8220 Compute Nodes, each with dual Intel Xeon E5-2670 (2.60GHz) 8 core CPUs, with 64 GB RAM. (Total Cores 528 and 1056 total threads, RAM$>$2TB). Source codes for 
SOAPdenovo2\footnote{SOAPdenovo2: http://soap.genomics.org.cn/soapdenovo.html} \cite{luo12}, UCLUST\footnote{UCLUST: http://www.drive5.com/uclust/downloads1\_{}2\_{}22q.html}    \cite{Edgar10}, and svm-light\footnote{svm-light: http://svmlight.joachims.org/} \cite{joachims08}
their respective websites and compiled on the Argo platform.

>>>>

We used data from a well-known Metagenome-Wide Association Study by Qin et al. of Type 2 Diabetes (T2D) in Chinese patients \cite{qin041012}. This study was chosen because it is one of the only MGWAS studies that made its data available online and labeled the phenotype of the patients, and is the largest among those studies. The full dataset used in this study contains 368 patients \cite{qin041012}, but only 218 of them were labeled and available online. Each patient file was downloaded from NCBI\footnote{http://trace.ncbi.nlm.nih.gov/Traces/study/?acc=SRP011011}
and converted to FASTQ format using the SRA toolkit\footnote{http://www.ncbi.nlm.nih.gov/Traces/sra/sra.cgi?view=software}.
The labels were found on the EBI website\footnote{https://www.ebi.ac.uk/metagenomics/projects/SRP011011}. The total size of these 218 FASTQ files was 2.3 terabytes, with an average size of 10.82 gigabytes per patient file. Qin et al. develop a simple T2D classifier using a minimum redundancy---maximum relevance (mRMR) method \cite{peng05} for feature selection and an SVM for classification, based on the R packages ``sideChannelAttack" and ``e1071", respectively \cite{qin041012}. They achieved an AUC-ROC of 0.81 by training a classifier on 345 of the patients and using the remaining 23 as a test set \cite{qin041012}. While we did not have access to all of the same patients that they had, we were able to replicate their method and test it on the subset of data that was publicly available. The study also called for more extensive testing of gut microbiota classifiers \cite{qin041012}.

\subsection{Software and Hardware Details}
We used the ARGO computing cluster available at George Mason University\footnote{http://orc.gmu.edu/research-computing/argo-cluster/argo-hardware-specs/}. The clustering and classification phases were run on one of the compute nodes available on the cluster. The cluster is configured with 35 Dell C8220 Compute Nodes, each with dual Intel Xeon E5-2670 (2.60GHz) 8 core CPUs, with 64 GB RAM. (Total Cores 528 and 1056 total threads, RAM$>$2TB). Source codes for 
SOAPdenovo2\footnote{SOAPdenovo2: http://soap.genomics.org.cn/soapdenovo.html} \cite{luo12}, UCLUST\footnote{UCLUST: http://www.drive5.com/uclust/downloads1\_{}2\_{}22q.html}    \cite{Edgar10}, and svm-light\footnote{svm-light: http://svmlight.joachims.org/} \cite{joachims08}
their respective websites and compiled on the ARGO platform.

### 2016-08-16 latex/input/methods.tex
finished related work section of paper and added diagrams; began working on results section
paper draft v1.1: most minor edits completed

As mentioned in the Introduction, our pipeline involves a number of steps, which serve a variety of purposes. For each patient file, we assembled the sequence reads, which served the dual purpose of generating larger contigs that contain more functional biological information and reducing the dataset size by discarding reads that could not be assembled. The reads that could not be assembled were unlikely to be useful anyway, since they could not be assembled with other reads into useful contigs. The clustering step assigns the contigs to certain clusters, which represent functionally similar microbes, and thus establish classes of instances that can be used as features for the classifier. We  applied our feature extraction method, which uses the D-BoW method of measuring the closest distance from a patient's reads to each cluster center. Finally, we used a standard SVM classifier to predict patient phenotype, and used several metrics to assess its accuracy. We used the SVM's decision boundary to infer information about which clusters of instances were most or least indicative of the phenotype, discussed further in the results section. Aside from the patient labels, this process is entirely de novo, and does not consult any external databases. An illustration of this pipeline can be seen in Figure \ref{pipeline} on page \pageref{pipeline}.

@@

For our assembly step, we used SOAPdenovo2 \cite{luo12}, which is a de novo metagenome assembly tool. This tool was also used for assembly in the MGWAS study that our data came from. We use this method in order to compare ourselves with the MGWAS study, and because Soapdenovo2 has been proven to be one of the fastest assemblers\cite{peng12}. We tested a number of different combinations of parameters, and found that the best results came when we cut reads off after 100 base pairs (reads were 180 base pairs long originally) and used a k-mer size of 51. The patient files needed to be assembled separately, in order to avoid assembling reads from different patients together. Conversely, all contigs need to be in one file for clustering, to avoid inconsistent cluster assignments between different patients. Thus, we combined the contigs from each assembled patient file into a single for clustering.

@@

We used UCLUST \cite{Edgar10}, which has been shown to be among the simplest, fastest, and most effective clustering methods \cite{bonder090112, sun042711}. Since our contigs were not ordered, we used the usersort option, and we set the sequence match id to 40\%, which means that two reads needed to have 40\% of the same nucleotides to be in the same cluster. For instance, between two strings of length 100, at least 40 places in each of those strings would have to contain the same nucleotide (represented as A, T, G, or C). New contigs that did not match at least 40\% to any of the existing contigs would form the seed of a new cluster. Again, this value of 40\% was found to be the best value based on our experiments with this dataset.

@@

Our feature extraction operation followed the D-BoW method. For each patient, we initialized a vector with length equal to the number of clusters. The value for each place in that patient's feature vector would then be set to the closest percentage match (0.4 to 1) of the reads from that patient to the relevant cluster seed. For instance, if there are three clusters of reads, and patient A has the cluster seed for cluster 1, no reads from cluster 2, and a read with 70\% match to cluster 3, their feature vector would be~~A = [1.0~~~0.0~~~0.7]. This is illustrated in Figure \ref{pipeline}.

>>>>

Our proposed pipeline involves a number of steps, which serve a variety of purposes. For each patient file, we assembled the sequence reads, which served the dual purpose of generating larger contigs that contain more functional biological information and reducing the dataset size by discarding reads that could not be assembled. The clustering step assigns the contigs to certain clusters, which represent functionally similar microbes, and thus establish classes of instances that can be used as features for the classifier. We  applied our feature extraction method, which uses the D-BoW method of measuring the closest distance from a patient's reads to each cluster center. Finally, we used an SVM-based classifier to predict patient phenotype, and used several metrics to assess its accuracy. We used the SVM's decision boundary to infer information about which clusters of instances were most or least indicative of the phenotype, discussed further in the Results section. Aside from the patient labels, this process is entirely de novo, and does not consult any external databases. An illustration of this pipeline can be seen in Figure \ref{pipeline} on page \pageref{pipeline}.

@@

For our assembly step, we used SOAPdenovo2 \cite{luo12}, which is a de novo metagenome assembly tool. This tool was also used for assembly in the MGWAS study that our data came from. We use this method in order to compare ourselves with the MGWAS study, and because Soapdenovo2 has been proven to be one of the fastest assemblers \cite{peng12}. We tested a number of different combinations of parameters, and found that the best results came when we cut reads off after 100 base pairs (reads were 180 base pairs long originally) and used a k-mer size of 51. The average insert size was set to 350, in accordance with the reported average insert size from the MGWAS study that we used data from \cite{qin041012}. The patient files needed to be assembled separately, in order to avoid assembling reads from different patients together. Conversely, all contigs need to be in one file for clustering, to avoid inconsistent cluster assignments between different patients. Thus, we combined the contigs from each assembled patient file into a single for clustering.

@@

We used UCLUST \cite{Edgar10}, which has been shown to be among the simplest, fastest, and most effective clustering methods \cite{bonder090112, sun042711}. Since our contigs were not ordered, we used the usersort option, and we set the sequence match threshold to 40\%, which means that two reads needed to have 40\% of the same nucleotides to be in the same cluster. For instance, between two strings of length 100, at least 40 places in each of those strings would have to contain the same nucleotide (represented as A, T, G, or C). New contigs that did not match at least 40\% to any of the existing contigs would form the seed of a new cluster. This value of 40\% was found to be the best value based on our experiments with this dataset. Other values tried, including 50\%, 75\%, and 90\%, led to many clusters with very few reads per cluster.

@@

Our feature extraction operation followed the D-BoW method. For each patient, we initialized a vector with length equal to the number of clusters. The value for each place in that patient's feature vector would then be set to the closest percentage match (0.4 to 1) of the reads from that patient to the relevant cluster seed. For instance, if there are three clusters of reads, and patient A has the cluster seed for cluster 1, no reads from cluster 2, and a read with 70\% match to cluster 3, their feature vector would be~~A = [1.0~~~0.0~~~0.7]. This is illustrated in Figure \ref{pipeline}. We implemented this feature extraction method in Python.
